Master's Research Proposal (IN PROGRESS)

### 2024-06-05 proposal/Proposal.tex
starter chinook eco

\subsection{Chinook Ecology}

\subsection{Chinook Management}

@@

>>>>

@@

\subsection{Chinook}

Chinook salmon (\textit{Oncorhynchus tshawytscha}) is the largest species of Pacific salmon. Like other Pacific salmon the species is both anadromous and semelparous leading to a very interesting and distinct life history.

Juvenile Chinook are born in rivers where they remain for some time before heading to coastal waters. When they make this migration down river is quite variable and dependent on the region in question with juveniles entering anywhere from their first to second years of life. This leads to a specific system of determining age that includes both years in freshwater and years at sea. For example if a particular fish spends one winter in freshwater and so far 2 at sea it would be a 1.2 age fish. If it spent zero winters in freshwater and 3 at sea it would be a 0.3 age fish. Clearly both of these examples are the same overall age.

The immature fish then spend time (up to five years) growing in the marine environment before maturing and then migrating back to the river where they were born to breed and subsequently die. The timing of such homing behavior is clearly temperature dependent with species further north having to avoid ice build up during the cooler months. The mechanism behind this homing ability is still a question under active research. \newline

Given this homing behavior Chinook populations are in fact a series of reasonably separated genetic sub-populations each of which can have its own life history strategies. Indeed using tagging data it has been demonstrated that both initial and later marine distribution is highly dependent on the river of origin for the fish in question.

Beyond this large scale analysis of fisheries dependent data has also demonstrated some of the likely migration patterns the species exhibits during its rearing phase at sea and electronic tagging studies have shown how quickly and with what accuracy natal homing can occur.

All in all then, Chinook represent a highly mobile species of fish.\newline

Broadly distributed through out the Northern pacific, chinook are known to spawn in rivers spanning from the Sacramento River in central California to the Kotzebue Sound in Alaska on the eastern side of the pacific and primarily in the Kamchatka Peninsula on the western side.

Marine residency in the North Eastern pacific is primarily coastal from California up to Western Alaska and then spans into much of the Eastern Bering Sea. \newline

Chinook are well known piscivores with analysis of stomach contents finding primarily fish and squid as their main sources of prey. It has been noted that this diet is similar to that of Coho salmon but that it is unlikely that the two species experience much cross-species competition for these resources.

Chinook themselves are prey for a few species such as Salmon sharks and whales and dolphins but it is still not clear how much mortality in later life stages is due to predation as opposed to other sources. Size selective predation on earlier life stages is a well known driver of natural mortality.

Other drivers of natural mortality include a variety of known diseases, starvation during low productivity months, and potential competition for food with species like whales.\newline

Chinook have been the subject of hatchery programs since the 1950s with the past decade seeing hatchery releases of 254million individuals on average. The vast majority of these individuals (~200million) have come from hatchery programs in the United States with the remaining significant portion from programs in Canada.

Fish are released at weights of around 6-9 grams and a large number of these have a micro-wire with an identifiable code etched into them inserted into their snouts. These are the tags driving the Coded Wire Tag program in the Northeastern Pacific that has become a major source of demographic and movement data used in managing and understanding Chinook sub-populations. The etchings are such that each cohort from each river can be identified allowing for long term studies of fish movement.

The initial purpose of these hatcheries programs was to supplement losses in juveniles due to targeting adult fish in river systems however the programs have not created the kind of recovery that was intended and it is unclear how wild and hatchery populations interact. For example it is well known that there is quite a bit of genetic exchange between wild and hatchery types and concern that hatchery types are less suited to wild environments. \newline

There are valuable tribal, commercial, and recreational fisheries of Chinook salmon through the Pacific.

Given the highly mobile nature of Chinook independent management by sovereign nations is simply not possible and so management is the ultimate responsibility of the Pacific Fishery Management Council. Federal management councils (NOAA for the United States) then collaborate with the Pacific Fishery Management Council to set harvest quotas and other management objectives. The governments of Cananda and the United States then work with states, tribes, and industry to create and manage shared objectives based on these overall goals.

These management goals as well as designations of statuses such as "overfished" have to be stated on the sub-population level given the fact that Chinook form so many reasonably separated genetic populations. Therefore management is further complicated by the fact that some sub populations are considered overfished while others are not. Indeed several sub populations are actually listed as endangered under the Endangered species act.

This latter fact as serious implications for non-Chinook fisheries such as the incredibly lucrative Walleye Pollock fishery which, if enough endangered Chinook are caught, can experience a moratorium. Therefore there is a great deal of interest in limiting bycatch of Chinook in these fisheries.

To manage this diversity the United States uses a variety of tools including but not limited to - differentiated management zones, minimum size limits, bag limits for recreational fisheries, allowed seasons and quotas, and fishing gear restrictions. \newline

Finally, Chinook salmon are especially prized for their large size which has led to serious questions about size selective mortality from fishing as both recreational and commercial fisheries have incentives for targeting larger fish. While it is not yet known whether fisheries are responsible there is clear evidence that fish size in Chinook populations has been decreasing steadily over the years which has potential impacts on both the fisheries themselves (in terms of optimizing yields) and for Chinook's primary predators. \newline

All in all then given their ecology and the corresponding complexities of management, Chinook represent an ideal model species for trying to understand and quantify spatial patterns of movement and behavior.
